---
title: "canaryGO_preprocessing"
author: "Carson Stacy"
date: "2022-07-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# load in annotations file file
canary_annotations <- readr::read_delim("~/Documents/GitHub/CanarySeq/geneAnnotations/eggNOG-mapper-v2/canary.annotations.tsv", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE, skip = 4) %>%
  dplyr::mutate(geneID_ncbi = stringr::str_sub(`#query`, start=24, end= 37))#make new column that is just the ncbi geneID

# make sure it looks pretty
View(canary_annotations)
```

```{r}
geneID2GO_build <- tibble(
                          # canary_annotations$geneID_ncbi,
                          canary_annotations$Preferred_name,
                          "\t",
                          canary_annotations$GOs
                          )
# write out the geneID2GO.map file
write.table(geneID2GO_build, file="~/Desktop/geneID2GO.map",
            quote = F, row.names = F, col.names = F)

# read in with readMappings
canary_geneID2GO <- topGO::readMappings(file = "~/Desktop/geneid2GO.map")

#remove genes that don't have a gene name from eggNOG output
canary_geneID2GO_filtered <- canary_geneID2GO[canary_geneID2GO != "-"]

# save these filtered mappings
save(canary_geneID2GO_filtered, file = "~/Desktop/canary/canaryGeneID2GOMappings.RData")
```


I want to combine gene names from eggNOG-mapper output with gene names from EdgeR output
```{r}
# load in edgeR output
MGLpost0 <- readr::read_csv("~/Desktop/canary/contrasts/MGLpost0.csv")

#load in canary geneID2GO map
load(file = "~/Desktop/canary/canaryGeneID2GOMappings.RData")

# Numeric ID values (in edgeR output) of genes whose names match Gene Ontology Estimates with edgeR output gene names
matches <- which(toupper(names(canary_geneID2GO_filtered)) %in% MGLpost0$...1)
          # these names came from `edgeR diff expression.R` file

# Numeric ID values (in eggNOG output) of genes whose names match Gene Ontology Estimates with edgeR output gene names
matches_rnaIndex <- which(MGLpost0$...1 %in%  toupper(names(canary_geneID2GO_filtered)))

```

```{r}
# eggNOG genes that didn't match
noMatch_eggNOG <- toupper(names(canary_geneID2GO_filtered))[-matches]

noMatch_edgeR <- MGLpost0$...1[-matches_rnaIndex]
```

```{r}
#genes that are in both.
# used unique because there are some duplicates in eggNOG output. (per isoform)
genesMatched <- unique(toupper(names(canary_geneID2GO_filtered))[matches])

# Identify genes names that are DE in edgeR with FDR < 0.05
# total of 273 first time I ran this with default settings
DEgenes <- filter(MGLpost0, FDR < 0.05) %>% pull(`...1`)

#count how many of the genes that are DE have GO terms.
# 187
length(which(DEgenes %in% genesMatched))

# genes names with GO terms and DE.
DEgeneswithGO <- which(DEgenes %in% genesMatched)

# select only genes that are DE and have GO terms
DEgenesGO <- DEgenes[DEgeneswithGO]
```
Here we see only 187 of the 273 $(68%%)$ DE genes in condition MGLpost0. It looks like many genes that don't have a "name" in the canary genome don't get matched to a gene with GO's attached. Not sure the best solution for this. Maybe trying InterProScan.

For current analysis, let me try doing GO analysis with just these genes. will figure out rest later.

```{r}
#first, let's subset our canary geneIDtoGO to just these genes.
canary_geneID2GO_filtered_DE <- #unique(
  canary_geneID2GO_filtered[which(toupper(names(canary_geneID2GO_filtered)) %in% DEgenesGO)] 
 # )
canary_geneID2GO_filtered_DE <- canary_geneID2GO_filtered_DE[!duplicated(canary_geneID2GO_filtered_DE)]

# it is weird there is a difference between length of these two when there shouldn't be. 
# Here are the two that are make canary_geneID2GO_filtered_DE longer than DEgenesGO
names(canary_geneID2GO_filtered_DE)[!(DEgenesGO %in% names(canary_geneID2GO_filtered_DE))]

# weird formatting creating the geneList for topGO. this code seemed to work for me.
all_genes <- factor(as.integer(toupper(MGLpost0$...1) %in% DEgenesGO)) # create 0 1 factors.
names(all_genes) <- toupper(MGLpost0$...1) # add gene names
str(all_genes)
```

```{r}
#create topGO data obj.
GOdata=new('topGOdata', 
           ontology='BP', # 'BP' = Biological Process. Also tried 'ALL' but doesn't work.
           allGenes = all_genes, # all 10188 genes that are in MGLpost0 data from edgeR. factor=1 for those that are DE.
           annot = annFUN.gene2GO, # annotation function for custom annotation
           gene2GO = canary_geneID2GO_filtered # annotation db from eggNOG output.
           # gene2GO = canary_geneID2GO_filtered_DE#, 
           # geneSel = DE_genes
           )
```

```{r}
# define test using the classic algorithm with fisher 
#(refer to Alexa,A., Rahnenführer,J. and Lengauer,T. (2006) Improved scoring of functional groups from gene expression data by decorrelating GO graph structure. Bioinformatics, 22, 1600–1607.
# for how the different algorithms work)
classic_fisher_result=runTest(GOdata, algorithm='classic', statistic='fisher')
```

```{r}
# define test using the weight01 algorithm (default) with fisher
weight_fisher_result=runTest(GOdata, algorithm='weight01', statistic='fisher') 
 
# generate a table of results: we can use the GenTable function to generate a summary table with the results from tests applied to the topGOdata object.
allGO=usedGO(GOdata)
all_res=GenTable(GOdata, weightFisher=weight_fisher_result, orderBy='weightFisher', topNodes=length(allGO))

#performing BH correction on our p values
p.adj=round(p.adjust(all_res$weightFisher,method="BH"),digits = 4)
 
# create the file with all the statistics from GO analysis
all_res_final=cbind(all_res,p.adj)
all_res_final=all_res_final[order(all_res_final$p.adj),]
 
#get list of significant GO before multiple testing correction
results.table.p= all_res_final[which(all_res_final$weightFisher<=0.05),]
 
#get list of significant GO after multiple testing correction
results.table.bh=all_res_final[which(all_res_final$p.adj<=0.05),]
 
#save first top 50 ontolgies sorted by adjusted pvalues
write.table(all_res_final[1:50,],"summary_topGO_analysis.csv",sep=",",quote=FALSE,row.names=FALSE)
 
# PLOT the GO hierarchy plot: the enriched GO terms are colored in yellow/red according to significance level
 
pdf(file='~/Desktop/canary/topGOPlot_fullnames.pdf', height=12, width=12, paper='special', pointsize=18)
showSigOfNodes(GOdata, score(weight_fisher_result), useInfo = "none", sigForAll=FALSE, firstSigNodes=2,.NO.CHAR=50)
dev.off()
```